Figure 3 - ChIP

### 2024-07-15 ms_figure_2_overview-heatmaps.Rmd
move RNA-level violin plot from fig3 report to suppFig2

---
title: "Figure 2 - Overview heatmaps for IP-MS and ChIP"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    complexes: "data/complexes.json"
    idmap: "data/id_mapping_table.txt"
    baitclass: "data/ipms_bait_class.txt"
    dbdtxt: "data/TF_DBD_annotation.txt"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakatacfile: "data/fused_peaks_filtered_counts-external_atac.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    ipms150untag: "data/ipms_150_untag_sce.rds"
---

@@

@@

## IP-MS data for HOT TFs

@@

# create main heatmap with annotations
hm3 <- Heatmap(m[sel_peaksBin, ],
               column_title = "ChIP (binary)",
               column_title_side = "top",
               column_title_gp = gpar(fontsize = ft),
               col = circlize::colorRamp2(
                   breaks = seq(0, max(m), length.out = 64),
                   colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               cluster_rows = TRUE, row_dend_width = unit(25, "mm"), show_row_dend = FALSE,
               cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksBin], cluster_row_slices = FALSE,
               left_annotation = peakAnnot,
               right_annotation = fracTFsAnnot,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = FALSE,
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm")) # whole heatmap height

@@

# create main heatmap with annotations
hm4 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
               column_title = "ChIP (enrichments)",
               column_title_side = "top",
               column_title_gp = gpar(fontsize = ft),
               col = circlize::colorRamp2(
                   breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                   colors = colorRampPalette(enrichment_heatmap_colors)(64)),
               cluster_rows = FALSE,
               cluster_columns = column_dend(hm3), column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksEnr], cluster_row_slices = FALSE,
               border = TRUE, border_gp = gpar(lwd = 0.5),
               left_annotation = peakAnnot,
               right_annotation = peakAnnot2,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = TRUE,
               heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                           labels_gp = gpar(fontsize = fs),
                                           border = "gray10",
                                           legend_direction = "horizontal",
                                           legend_width = unit(30, "mm"),
                                           title = "TF log2 IP/input",
                                           title_position = "topcenter",
                                           title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm"))

@@

## ChIP enrichment replicate-pair correlations

@@

@@

cowplot::plot_grid(
    gg_enrpairs,
    hm_ipms_hot,
    rel_widths = c(1, 0.65), 
    scale = 0.9, nrow = 1,
    labels = c("A", "B")
)
```

>>>>

---
title: "Figure 2 - Overview heatmaps for IP-MS and ChIP"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    complexes: "data/complexes.json"
    idmap: "data/id_mapping_table.txt"
    baitclass: "data/ipms_bait_class.txt"
    dbdtxt: "data/TF_DBD_annotation.txt"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakatacfile: "data/fused_peaks_filtered_counts-external_atac.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    rdsrnaseq_del: "data/rnaseq_del.rds"
    txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    ipms150untag: "data/ipms_150_untag_sce.rds"
---

@@

# `rnaseq_del`: a `SummarizedExperiment` with RNA-seq quantifications
rnaseq_del <- readRDS(params$rdsrnaseq_del)

# `txannot`: data.frame with transcript start site (TSS) and
#            transcript end site (TES) annotation
txannot <- read.csv(params$txcsv, row.names = 1)

@@

## Supplementary plots - IP-MS data for HOT TFs

@@

# create main heatmap with annotations
hm3 <- Heatmap(m[sel_peaksBin, ],
               column_title = "Binarized ChIP enrichments",
               column_title_side = "top",
               column_title_gp = gpar(fontsize = ft),
               col = circlize::colorRamp2(
                   breaks = seq(0, max(m), length.out = 64),
                   colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               cluster_rows = TRUE, row_dend_width = unit(25, "mm"), show_row_dend = FALSE,
               cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksBin], cluster_row_slices = FALSE,
               left_annotation = peakAnnot,
               right_annotation = fracTFsAnnot,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = FALSE,
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm")) # whole heatmap height

@@

# create main heatmap with annotations
hm4 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
               column_title = "ChIP enrichments",
               column_title_side = "top",
               column_title_gp = gpar(fontsize = ft),
               col = circlize::colorRamp2(
                   breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                   colors = colorRampPalette(enrichment_heatmap_colors)(64)),
               cluster_rows = FALSE,
               cluster_columns = column_dend(hm3), column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksEnr], cluster_row_slices = FALSE,
               border = TRUE, border_gp = gpar(lwd = 0.5),
               left_annotation = peakAnnot,
               right_annotation = peakAnnot2,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = TRUE,
               heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                           labels_gp = gpar(fontsize = fs),
                                           border = "gray10",
                                           legend_direction = "horizontal",
                                           legend_width = unit(30, "mm"),
                                           title = "TF log2 IP/input",
                                           title_position = "topcenter",
                                           title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm"))

@@

## Supplementary plots - ChIP enrichment replicate-pair correlations

@@

## Supplementary plots - Bound promoters

In `txannot$prom.number.overlapping.highconf.peaks`, we record how many TF peaks
are found in the promoter regions of a gene (defined as a 1kb window
centered on the annotated transcript start site).

Here we compare the expression levels (obtained from `rnaseq_del`) of
genes twith different numbers of TFs bound at their promoters.

```{r bound-promoters}
#| fig.width: 6
#| fig.height: 4

# calculate average expression of each gene
avg_tpm <- rowMeans(assay(rnaseq_del, "abundance"))

# add average expression to gene annotation and filter out
# genes that are not protein coding genes
pd <- txannot |>
    mutate(avg_tpm = avg_tpm[unique_einprot_id]) |>
    filter(!is.na(avg_tpm),
           gene_biotype == "protein_coding") |>
    mutate(prom.number.overlapping.highconf.peaks.categories = 
               cut(prom.number.overlapping.highconf.peaks,
                   breaks = c(-Inf, 0, 1, Inf),
                   labels = c("0", "1", ">1"),
                   right = TRUE))

# visualize
(gg_rnaviolin <- ggplot(pd, aes(prom.number.overlapping.highconf.peaks.categories, avg_tpm + 0.01)) +
        geom_violin(draw_quantiles = 0.5, fill = bg_color) +
        scale_y_log10(labels = scales::label_number(big.mark = "")) +
        labs(x = "Number of promoter TF peaks",
             y = "RNA level\n(TPM + 0.01, log-scale)") +
        theme_cowplot())
```

@@

cowplot::plot_grid(
    gg_enrpairs,
    cowplot::plot_grid(
        hm_ipms_hot,
        gg_rnaviolin,
        NULL,
        NULL,
        ncol = 1,
        rel_heights = c(1.25, 0.75, 1, 1),
        labels = c("", "C", "D", "E")
    ),
    rel_widths = c(1, 0.65), 
    scale = 0.9, nrow = 1,
    labels = c("A", "B")
)
```

### 2024-07-15 ms_figure_3_chip.Rmd
move RNA-level violin plot from fig3 report to suppFig2

---
title: "Figure 3 - ChIP"
author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    genomefasta: "reference/Schizosaccharomyces_pombe.ASM294v2.dna.toplevel.fa"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    chiptrackfile: "data/genome_browser_tracks_chipseq.csv.gz"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    dbdtxt: "data/TF_DBD_annotation.txt"
    motifrds: "data/motifs_annotated.rds"
    rdsrnaseq_del: "data/rnaseq_del.rds"
    txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
editor_options: 
    chunk_output_type: console
---

@@

# `rnaseq_del`: a `SummarizedExperiment` with RNA-seq quantifications
rnaseq_del <- readRDS(params$rdsrnaseq_del)

@@

## Bound promoters

In `txannot$prom.number.overlapping.highconf.peaks`, we record how many TF peaks
are found in the promoter regions of a gene (defined as a 1kb window
centered on the annotated transcript start site).

Here we compare the expression levels (obtained from `rnaseq_del`) of
genes twith different numbers of TFs bound at their promoters.

```{r bound-promoters}
#| fig.width: 6
#| fig.height: 4

# calculate average expression of each gene
avg_tpm <- rowMeans(assay(rnaseq_del, "abundance"))

# add average expression to gene annotation and filter out
# genes that are not protein coding genes
pd <- txannot |>
    mutate(avg_tpm = avg_tpm[unique_einprot_id]) |>
    filter(!is.na(avg_tpm),
           gene_biotype == "protein_coding") |>
    mutate(prom.number.overlapping.highconf.peaks.categories = 
               cut(prom.number.overlapping.highconf.peaks,
                   breaks = c(-Inf, 0, 1, Inf),
                   labels = c("0", "1", ">1"),
                   right = TRUE))

# visualize
ggplot(pd, aes(prom.number.overlapping.highconf.peaks.categories, avg_tpm + 0.01)) +
    geom_violin(draw_quantiles = 0.5) +
    scale_y_log10() +
    labs(x = "Number of promoter TF peaks",
         y = "RNA level (TPM + 0.01, log-scale)") +
    theme_cowplot()
```

>>>>

---
title: "Figure 3 - ChIP"
author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    genomefasta: "reference/Schizosaccharomyces_pombe.ASM294v2.dna.toplevel.fa"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    chiptrackfile: "data/genome_browser_tracks_chipseq.csv.gz"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    dbdtxt: "data/TF_DBD_annotation.txt"
    motifrds: "data/motifs_annotated.rds"
    txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
editor_options: 
    chunk_output_type: console
---

@@

@@
